---
title: "Project title"
author: "Minh Tran,"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE, message = FALSE, warning = FALSE}

library(mosaic)
library(tidyverse)
library(survey)
library(knitr)

```

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```


# Abstract
## One paragraph summary of the purpose, dataset, main findings, and interpretation.

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

# Introduction

## Brief description of the dataset (subjects, when/where/how collected, purpose).

## Research questions

## Mini codebook of variables selected (include details on recoding values, missing values, and variable treatment)

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

# Statistical Results

## One subsection per research question.

## Include summary statistics, tables, graphs, and statistical tests.

## Write a clear narrative (“tell the story of your data”)

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

# Conclusion and Future Direction

## Summarize main findings.

## Discuss limitations of your dataset.

## Suggest possible extensions or improvements.

```{r, results='asis', eval=(opts_knit$get('rmarkdown.pandoc.to') == 'latex'), echo = FALSE}
cat('\\pagebreak')
```

# Appendix (R code)




